\ZpL: a p-adic precision package

### 2018-02-02 ZpL/ZpL.tex

\begin{itemize}
\item Interval arithmetic, floating point arithmetic
\item short introduction of the precision lemma and lattices
\item goal: tracking precision using this theory
\item short presentation of the package
\item plan of the article
\end{itemize}

@@

In \cite{LV16}, the behaviour of the precision when solving
$p$-adic differential equations with separation of 
variables has been studied.
The authors have investigated the gap when that appears
when applying a Newton-method solver between
the theoretic loss in precision and the 
actual loss in precision for a naive implemenationin \verb?Zp(p)?.
We can reach this theoretical loss in precision using \verb?ZpL?.
We use a generic {\color{function}\verb?Newton_Iteration_Solver?}\verb?(g,h,N)?
that applies \verb?N? steps of the Newton method for 
$y'=g \times h(y)$ as described in \citep{LV16}.

@@

For instance, the curves 
$E: \: y^2=x^3+1$
and $E': \: y^2=+ (m^2-1)x^2/4 + m^6$ are isogeneous
over $\mathbb{Q}$ if  $m$ is an odd integer.
We take $m=67.$ The corresponding isogeny has coefficients with possibly
negative $3$-adic valuation.

>>>>

When computing with real and $p$-adic fields, exact results are usually impossible, since most elements have infinite decimal or $p$-adic expansions.  Working with these fields thus requires an analysis of how precision evolves through the sequence of steps involved in carrying out a computation.  In this paper, we describe a package for computing with $p$-adic rings and fields, based on a series of papers by the same authors \todo{include citations}.  The core of the package is a method for tracking precision using $p$-adic lattices which can yield dramatically more precise results, at the cost of increased runtime and memory usage.

The standard method for handling precision when computing with real numbers is floating point arithmetic, which may also be used in $p$-adic computation.  At a given precision level, a finite set of representable numbers are chosen, and arithmetic operations are defined to give a representable number that is close to the true result \todo{add reference to floating point arithmetic}.  For example, the representable numbers in $53$-bit floating point arithmetic are those of the form $m \cdot 2^e$ \todo{describe valid $m$ and $e$}.  Floating point arithmetic has the benefit of efficient arithmetic operations, but users are responsible for tracking the precision of the results.  Numerically unstable algorithms can lead to very inaccurate answers \todo{add reference for numerical stability}.

If provably correct results are desired, interval arithmetic provides an alternative to floating point.  Instead of just tracking an approximation to the answer, the package also tracks a radius within which the true result lies.  This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.  Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision \todo{include reference to SOMOS section}.

For better control of precision, we may use ellipsoids rather than balls.  This change would have little utility for real numbers, since ellipsoids are not preserved by most functions.  For $p$-adic fields, in contrast, sufficiently small ellipsoids are preserved by differentiable maps.  From an algebraic perspective, these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space, and Lemma \todo{add number} describes how the image of such a coset under a map $f$ is given exactly by applying the differential of $f$ to $H$.

In this paper, we describe an implementation of this idea in Sage \todo{cite Sage}.  \todo{Add short description of package features.}

\todo{Plan of article}

@@

In \cite{LV16}, the behaviour of the precision when solving
$p$-adic differential equations with separation of 
variables has been studied.
The authors have investigated the gap when that appears
when applying a Newton-method solver between
the theoretic loss in precision and the 
actual loss in precision for a naive implemenationin \verb?Zp(p)?.
We can reach this theoretical loss in precision using \verb?ZpL?.
We use a generic {\color{function}\verb?Newton_Iteration_Solver?}\verb?(g,h,N)?
that applies \verb?N? steps of the Newton method for 
$y'=g \times h(y)$ as described in \cite{LV16}.

@@

For instance, the curves 
$E: \: y^2=x^3+1$
and $E': \: y^2= x^3+ (m^2-1)x^2/4 + m^6$ are isogeneous
over $\mathbb{Q}$ if  $m$ is an odd integer.
We take $m=67.$ The corresponding isogeny has coefficients with possibly
negative $3$-adic valuation.
